---
title: "Descriptive"
author: "Elia Gonzato"
date: "2024-05-28"
output: html_document
---

Load libraries

```{r, warning=F,message=F,echo=F}
library(pacman)
p_load(ggplot2,
       readxl,
       dplyr,
       labelled,
       gtsummary,
       ggalluvial)
```

Read clean dataset from folder and change values of `EQ5D5L` items.

```{r, warning=F,message=F,echo=F}
wave1234=read_xlsx('O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Clean dataset\\w1234.xlsx')
# apply same transformation to different variables
wave1234c=wave1234 %>% 
  #filter(Gender!=3) %>% 
  mutate_at(vars(MO,PD,UA,AD,SC),~case_when(.==1~'No Problems',
                                             .==2~'Slight Problems',
                                             .==3~'Moderate Problems',
                                             .==4~'Severe Problems',
                                             .==5~'Extreme Problems')) %>% 
  mutate(AgeCat=ifelse(is.na(AgeCat),'65-75',AgeCat)) 
# keep first row of all the waves for no response analysis
wave1t=wave1234c %>% 
  filter(t==1) %>% 
  mutate(Answer='Yes') %>% 
  select(pid,Answer)
```

## Non response analysis

Read `Wave13` dataset 

```{r}
wave13=haven::read_sav('O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\Wave 1 merged with 3_US_UK_NL_GR_IT_NY_SE___final working dataset_ES.sav')
# now clean to keep only the variables of interest
wave13c=wave13 %>% 
  mutate(Country=qCountry_w1,
         Gender=Gender_w1,
         Age=Age_w1,
         AgeCat=Age_cat_w1,
         MO=eq5d_mo_w1,
         SC=eq5d_sc_w1,
         UA=eq5d_ua_w1,
         PD=eq5d_pd_w1,
         AD=eq5d_ad_w1,
         VAS=eq_vas_w1,
         Comorb=chronic_none_w1,
         Covid=case_when(covid_status_w1== 1 ~ 1,
                         covid_status_w1 %in% c(2,3,4) ~ 2,
                         T ~ 3),
         Education=Education_cat_w1) %>% 
  select(pid,Country,Age,AgeCat,Gender,MO,SC,UA,PD,PD,AD,VAS,Comorb,Covid,Education)%>% 
  mutate(across(c(pid,Country,Age,AgeCat,Gender,MO,SC,UA,PD,AD,VAS),as.numeric))
```

Now merge the two datasets

```{r}
nradf=merge(wave13c,wave1t,by='pid',all.x=T) %>% 
  mutate(Answer=ifelse(is.na(Answer),'No',Answer),
         Education=case_when(Education==1~'Low',
                             Education==2~'Middle',
                             T~'High'),
         Comorb=ifelse(Comorb==0,'No','Yes'),
         Gender=ifelse(Gender==1,'Male','Female'),
         Covid=case_when(Covid==1~'No',
                         Covid==2~'Likely',
                         T~'Yes'))
# ref level
nradf$Answer=factor(nradf$Answer,levels=c('Yes','No'))
# now table demographics and other variables of interest by answer
nradf %>% 
  select(Answer,Age,Gender,Education,Comorb,Covid) %>% 
  tbl_summary(by=Answer) %>% 
  add_p() %>% 
  #bold_labels() %>% 
  #bold_p() %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\NonResponseAnalysis.xlsx')
```


## Descriptive statistics

Table characteristics at baseline:

```{r}
wavebase=wave1234c %>% 
  dplyr::filter(t==1) %>% 
  mutate(Education=case_when(Education==1~'Low',
                             Education==2~'Middle',
                             T~'High'),
         Comorb=ifelse(Comorb==0,'No','Yes'),
         Gender=ifelse(Gender==1,'Male','Female'),
         Covid=case_when(Covid==1~'No',
                         Covid==2~'Likely',
                         T~'Yes'))
# table Country, Age, Gender, Education, Covid, Comorb
wavebase %>% 
  select(Age, Gender, Education, Covid, Comorb,Country) %>% 
  tbl_summary(statistic=list( all_continuous() ~ "{median} ({p25}, {p75})")) %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\Table1.xlsx')
```

Go long to table once

```{r, warning=F,message=F,echo=F}
long=tidyr::gather(wave1234c,condition,value,c(MO,PD,UA,AD,SC)) %>% 
  mutate(condition=case_when(condition=='MO'~'Mobility',
                             condition=='PD'~'Pain Discomfort',
                             condition=='UA'~'Usual Activities',
                             condition=='AD'~'Anxiety Depression',
                             T~'Self-Care'))
```

Change values to variables of interest

```{r, warning=F,message=F,echo=F}
# create variables with the right name to display tables
wave1234t=wave1234c %>% 
  mutate(`Mobility`=MO,
         `Self-Care`=SC,
         `Usual Activities`=UA,
         `Pain Discomfort`=PD,
         `Anxiety Depression`=AD,
         Year=case_when(t==1~'2020',
                        t==2~'2021',
                        t==3~'2022',
                        T~'2023')) %>% 
  mutate_at(vars(`Mobility`,`Self-Care`,`Usual Activities`,
                 `Pain Discomfort`,`Anxiety Depression`),~factor(.,
                                                                 levels=c('No Problems',
                                                                          'Slight Problems',
                                                                          'Moderate Problems',
                                                                          'Severe Problems',
                                                                          'Extreme Problems'))) 
```

Table Age, Gender and Countries by year.

```{r, warning=F,message=F,echo=F}
wave1234t %>% 
  filter(Gender!=3) %>%
  mutate(Education=case_when(Education==1~'Low',
                             Education==2~'Middle',
                             T~'High'),
         Comorb=ifelse(Comorb==0,'No','Yes'),
         Gender=ifelse(Gender==1,'Male','Female'),
         Covid=case_when(Covid==1~'No',
                         Covid==2~'Likely',
                         T~'Yes')) %>% 
  select(Age, Gender, Education, Covid, Comorb,Country,Year) %>% 
  tbl_summary(by=Year,
              statistic=list( all_continuous() ~ "{median} ({p25},{p75})"),
              digits = list(all_continuous() ~ c(0, 0))) %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\TableYear.xlsx')
```


Distribution of Age

```{r, warning=F,message=F,echo=F}
wave1234t %>% 
  filter(t==4) %>% 
  ggplot(.,aes(x=Age))+
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue",color = "black") +
  geom_density(col = "navy") +
  labs(x = "Your Variable Name")+
  #geom_col(color = "black")+
  labs(x='Age',
       y='Frequency')+
  theme_bw()
```


Provide table that lists results of the five dimensions by time:

```{r, warning=F,message=F,echo=F}
# table univariate
wave1234t %>% 
  #filter(Year=='2020' | Year=='2023') %>% 
  select(`Mobility`,`Self-Care`,`Usual Activities`,`Pain Discomfort`,`Anxiety Depression`,Year) %>% 
  tbl_summary(by=Year) %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\TableDimensionsYear.xlsx')
```

Dimensions in the first year

```{r}
wave1234t %>% 
  filter(Year=='2020') %>% 
  select(`Mobility`,`Self-Care`,`Usual Activities`,`Pain Discomfort`,`Anxiety Depression`,Year) %>% 
  tbl_summary(by=Year) %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\TableDimensions2020.xlsx')
```

Table between year 2020 and year 2024

```{r}
wave1234t %>% 
  filter(Year=='2020' | Year=='2023') %>% 
  select(`Mobility`,`Self-Care`,`Usual Activities`,`Pain Discomfort`,`Anxiety Depression`,Year) %>% 
  tbl_summary(by=Year) %>% 
  add_p() %>% 
  bold_p() %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\TableDimensions2023.xlsx')
```


Now the same as before, but for the first year, and by gender

```{r, warning=F,message=F,echo=F}
wave1234t %>% 
  filter(Year=='2020' & Gender!=3) %>% 
  mutate(Gender=ifelse(Gender==1,'Male','Female')) %>% 
  select(`Mobility`,`Self-Care`,`Usual Activities`,`Pain Discomfort`,`Anxiety Depression`,
         Gender) %>% 
  tbl_summary(by=Gender) %>% 
  bold_labels() %>% 
  #add_p() %>% 
  #bold_p() %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\TableDimensionsGender2020.xlsx')
```

Now the same as before, but for the first year, and by gender

```{r, warning=F,message=F,echo=F}
wave1234t %>% 
  filter(Year=='2023' & Gender!=3) %>% 
  mutate(Gender=ifelse(Gender==1,'Male','Female')) %>% 
  select(`Mobility`,`Self-Care`,`Usual Activities`,`Pain Discomfort`,`Anxiety Depression`,
         Gender) %>% 
  tbl_summary(by=Gender) %>% 
  #bold_labels() %>% 
  add_p() %>% 
  bold_p() %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\TableDimensionsGender2023.xlsx')
```


Now the same as before, but for the first year, by Age Group

```{r, warning=F,message=F,echo=F}
wave1234t %>% 
  #mutate(AgeDummy=ifelse(AgeCat %in% c(1,2),'Before 2,included','After 2')) %>% 
  filter(Year=='2020') %>% 
  select(`Mobility`,`Self-Care`,`Usual Activities`,`Pain Discomfort`,`Anxiety Depression`,
         AgeCat) %>% 
  tbl_summary(by=AgeCat) %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\TableDimensionsAgeCat2020.xlsx')
```

Now the same as before, but for the first year, by Age Group

```{r, warning=F,message=F,echo=F}
wave1234t %>% 
  #mutate(AgeDummy=ifelse(AgeCat %in% c(1,2),'Before 2,included','After 2')) %>% 
  filter(Year=='2023') %>% 
  select(`Mobility`,`Self-Care`,`Usual Activities`,`Pain Discomfort`,`Anxiety Depression`,
         AgeCat) %>% 
  tbl_summary(by=AgeCat) %>% 
  gtsummary::as_tibble() %>% 
  writexl::write_xlsx(.,'O:\\PRJCT\\D&I\\3. Research Projects\\Ongoing projects\\POPCORN\\Elia\\All material\\Tables\\TableDimensionsAgeCat2023.xlsx')
```

Table conditions, by time

```{r, warning=F,message=F,echo=F}
stack=as.data.frame(table(long$value,long$t,long$condition))
names(stack)=c('Condition','Time','Dimension','Count')
stack=stack %>% 
  group_by(Time,Dimension) %>% 
  mutate(Sum=sum(Count),
         Freq=Count/Sum*100,
         Condition=factor(Condition,levels=c('Extreme Problems',
                                             'Severe Problems',
                                             'Moderate Problems',
                                             'Slight Problems',
                                             'No Problems')),
         Time=case_when(Time==1~'2020',
                        Time==2~'2021',
                        Time==3~'2022',
                        T~'2023'),
         Dimension=factor(Dimension,levels=(c('Mobility',
                                           'Self-Care',
                                           'Usual Activities',
                                           'Pain Discomfort',
                                           'Anxiety Depression'))))
```

Tables VAS Mean by year and test differences

```{r, warning=F,message=F,echo=F}
wave1234 %>% 
  mutate(Time=case_when(t==1~'2020',
                        t==2~'2021',
                        t==3~'2022',
                        T~'2023')) %>% 
  select(Time,VAS) %>% 
  tbl_summary(by=Time,
              statistic = list(all_continuous() ~ "{mean} ({sd})"),
              digits = list(VAS ~ c(1, 1))) %>% 
  add_p()
```


Tables VAS Mean by Country and test differences

```{r, warning=F,message=F,echo=F}
wave1234 %>% 
  mutate(Time=case_when(t==1~'2020',
                        t==2~'2021',
                        t==3~'2022',
                        T~'2023')) %>% 
  select(Country,VAS) %>% 
  tbl_summary(by=Country,
              statistic = list(all_continuous() ~ "{mean} ({sd})"),
              digits = list(VAS ~ c(1, 1))) %>% 
  add_p()
```

Now plot with `ggplot`

```{r, warning=F,message=F,echo=F}
ggplot(stack,aes(x=Time,y=Freq,fill=Condition))+
  geom_bar(stat='identity',width=0.5,col='black')+
  facet_wrap(~Dimension,nrow=1)+
  scale_fill_manual(values=c("#b3d9ff", "#66a3ff",
                             "#005c99", "#003366", "#001f33"),
                    breaks=c('No Problems','Slight Problems',
                             'Moderate Problems','Severe Problems',
                             'Extreme Problems'))+
  #scale_fill_discrete()+
  scale_color_manual(values = rep('black',5))+
  theme_bw()+
  theme(strip.text = element_text(face="bold", size=9),
        strip.background = element_rect(fill='white'),
        legend.position = 'bottom')+
  guides(fill = guide_legend(title.position="top", title.hjust = 0.5))+
  geom_col(color = "black")+
  labs(x='Year',
       y='Frequency') +
  theme(panel.grid = element_blank())
```

Now chart by Age, create table by age groups

```{r, warning=F,message=F,echo=F}
long.chart=long %>% 
  filter(t==3 & Gender !=3) %>% 
  mutate(Problems=ifelse(value=='No Problems',0,1),
         Sex=ifelse(Gender==1,'Male','Female'))
# now table Problems by condition,gender and age groups
chart=as.data.frame(table(long.chart$Problems,long.chart$AgeCat,
                          long.chart$Sex,long.chart$condition))
names(chart)=c('Problems','AgeCat','Gender','Dimension','Count')
# now wrangle to calculate proportions
chart=chart %>% 
  group_by(AgeCat,Gender,Dimension) %>% 
  mutate(Sum=sum(Count),
         Freq=Count/Sum,
         Dimension=factor(Dimension,levels=(c('Mobility',
                                           'Self-Care',
                                           'Usual Activities',
                                           'Pain Discomfort',
                                           'Anxiety Depression')))) %>% 
  filter(Problems==1)
```

Now plot results

```{r, warning=F,message=F,echo=F}
ggplot(chart,aes(x=AgeCat,y=Freq,color=Gender,group=Gender))+
  geom_point()+
  geom_line()+
  theme_bw()+
  facet_wrap(~Dimension,nrow=1)+
  theme(strip.text = element_text(face="bold", size=9),
        strip.background = element_rect(fill='white'),
        legend.position = 'bottom')+
  guides(fill = guide_legend(title.position="top", title.hjust = 0.5))+
  scale_color_manual(name='Gender',values = c("#FC0E07","#55C8F8"))+
  labs(x='Age Category',
       y='Frequency')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Mean of VAS by year and country

```{r, warning=F,message=F,echo=F}
means=wave1234 %>% 
  group_by(t,Country) %>% 
  summarize(vas_mean=mean(VAS)) %>% 
  mutate(t=case_when(t==1~2020,
                     t==2~2021,
                     t==3~2022,
                     T~2023))
# overall mean
mean_o=wave1234 %>% 
  group_by(t) %>% 
  summarize(vas_mean=mean(VAS)) %>% 
  mutate(t=case_when(t==1~2020,
                     t==2~2021,
                     t==3~2022,
                     T~2023),
         Country='Overall') %>% 
  select(t,Country,vas_mean)
# bind the two
#means=rbind(means,mean_o)
# plot
ggplot(means,aes(x=t,y=vas_mean,group=Country,color=Country))+
  geom_point()+
  geom_line()+
  geom_point(data=mean_o,aes(x=t,y=vas_mean),color='black')+
  geom_line(data=mean_o,aes(x=t,y=vas_mean),color='black')+
  theme_bw()+
  ylim(50,85)+
  labs(x='Years',
       y='VAS Mean')
``` 

## Mean of VAS by year and age groups

```{r, warning=F,message=F,echo=F}
means=wave1234 %>%
  filter(!(is.na(AgeCat))) %>% 
  mutate(Gender=factor(ifelse(Gender==1,'Male','Female'))) %>% 
  group_by(t,AgeCat,Gender) %>% 
  summarize(vas_mean=mean(VAS)) %>% 
  mutate(t=case_when(t==1~2020,
                     t==2~2021,
                     t==3~2022,
                     T~2023)) %>% 
  ungroup()
# now plot
ggplot(means,aes(x=AgeCat,y=vas_mean,group=Gender,color=Gender)) +
  geom_point()+
  geom_line()+
  facet_wrap(~t,nrow=1)+
  theme_bw()+
  scale_color_manual(name='Gender',values = c("#FC0E07","#55C8F8"))+
  labs(y='EQVAS Mean',x='Age Category')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  theme(strip.text = element_text(face="bold", size=9),
        strip.background = element_rect(fill='white'),
        legend.position = 'bottom')+
  guides(fill = guide_legend(title.position="top", title.hjust = 0.5))
```

